---
title: "Twang only analysis summary"
author: "Erik Westlund"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("settings.R")

library(dplyr)
library(kableExtra)
library(MatchIt)
library(twang)
library(survey)
library(margins)

# Note this depends on output from analysis.rmd

m_data <- readRDS(file.path(ps_dir, "m_data.rds"))
twang <- readRDS(file.path(ps_dir, "twang_gbm.rds"))
formulas <- readRDS(file.path(ps_dir, "formulas.rds"))
```

# Model

This model adjusts for multiple eyes by using model-robust standard errors (specifically, Horvitz-Thompson standard errors, which are generalized from model-robust sandwich estimators). Details are in the documentation for the survey package.

The below output is before exponentiation. To exponentiate any coefficient, just do `e^b` where `b` is the coefficient and `e` is Euler's constant. In `R` or `Stata`, type `exp(b)` for short-hand

```{r analysis}
m_data$twang_att_w <- get.weights(twang$result, stop.method = "es.mean")
design <- svydesign(ids = ~e_mrn_deidentified, weights = ~twang_att_w, data = m_data)
model <- svyglm(formula = formulas[["full"]], design = design, family = quasibinomial())

summary(model)
```

# Marginal effects

Inspect logic for bugs given quickness of this turnaround

```{r mfx}
avg_margfx <- margins(model, design=design)

all_mfx <- summary(avg_margfx) 

# PDR on any treatment
design_pdr_any_tx <- subset(design, PDR == "Present" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
data_pdr_any_tx <- subset(m_data, PDR == "Present" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
n_any_tx <- data_pdr_any_tx |> nrow()
mfx_any_tx <- summary(margins(model, design = design_pdr_any_tx, data = data_pdr_any_tx, variables=c("ever_lapse_binary")))

# PDR on anti-VEGF only
design_pdr_anti_vegf <- subset(design, PDR == "Present" & PRP_flag == 0 & anti_VEGF == 1 & focal_laser_flag == 0 & other_inject == 0)
data_pdr_anti_vegf <- subset(m_data, PDR == "Present" & PRP_flag == 0 & anti_VEGF == 1 & focal_laser_flag == 0 & other_inject == 0)
n_anti_vegf <- data_pdr_anti_vegf |> nrow()
mfx_anti_vegf <- summary(margins(model, design = design_pdr_anti_vegf, data = data_pdr_anti_vegf, variables=c("ever_lapse_binary")))

# PDR on PRP only
design_pdr_prp <- subset(design, PDR == "Present" & PRP_flag == 1 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0)
data_pdr_prp <- subset(m_data, PDR == "Present" & PRP_flag == 1 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0)
n_pdr_prp <- data_pdr_prp |> nrow()
mfx_pdr_prp <- summary(margins(model, design = design_pdr_prp, data = data_pdr_prp, variables=c("ever_lapse_binary")))

# NPDR or NoDR not on treatment
design_npdr_nodr_no_tx <- subset(design, (!is.na(NPDR) | No_DR == TRUE) & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
data_npdr_nodr_no_tx <- subset(m_data, (!is.na(NPDR) | No_DR == TRUE) & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
n_npdr_nodr_no_tx <- data_npdr_nodr_no_tx |> nrow()
mfx_npdr_nodr_no_tx <- summary(margins(model, design = design_npdr_nodr_no_tx, data = data_npdr_nodr_no_tx, variables=c("ever_lapse_binary")))

# NPDR or NoDR any tx
design_npdr_nodr_any_tx <- subset(design, (!is.na(NPDR) | No_DR == TRUE) & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
data_npdr_nodr_any_tx <- subset(m_data, (!is.na(NPDR) | No_DR == TRUE) & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
n_npdr_nodr_any_tx <- data_npdr_nodr_any_tx |> nrow()
mfx_npdr_nodr_any_tx <- summary(margins(model, design = design_npdr_nodr_any_tx, data = data_npdr_nodr_any_tx, variables=c("ever_lapse_binary")))

# NPDR not on treatment
design_npdr_no_tx <- subset(design, !is.na(NPDR) & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
data_npdr_no_tx <- subset(m_data, !is.na(NPDR) & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
n_npdr_no_tx <- data_npdr_no_tx |> nrow()
mfx_npdr_no_tx <- summary(margins(model, design = design_npdr_no_tx, data = data_npdr_no_tx, variables=c("ever_lapse_binary")))

# NPDR  on any treatment
design_npdr_any_tx <- subset(design, !is.na(NPDR) & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
data_npdr_any_tx <- subset(m_data, !is.na(NPDR) & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
n_npdr_any_tx <- data_npdr_any_tx |> nrow()
mfx_npdr_any_tx <- summary(margins(model, design = design_npdr_any_tx, data = data_npdr_any_tx, variables=c("ever_lapse_binary")))

# NPDR not on treatment: Mild
design_npdr_mild_no_tx <- subset(design, NPDR == "Mild" & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
data_npdr_mild_no_tx <- subset(m_data, NPDR == "Mild" & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
n_npdr_mild_no_tx <- data_npdr_mild_no_tx |> nrow()
mfx_npdr_mild_no_tx <- summary(margins(model, design = design_npdr_mild_no_tx, data = data_npdr_mild_no_tx, variables=c("ever_lapse_binary")))

# NPDR  on any treatment: Mild
design_npdr_mild_any_tx <- subset(design, NPDR == "Mild" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
data_npdr_mild_any_tx <- subset(m_data, NPDR == "Mild" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
n_npdr_mild_any_tx <- data_npdr_mild_any_tx |> nrow()
mfx_npdr_mild_any_tx <- summary(margins(model, design = design_npdr_mild_any_tx, data = data_npdr_mild_any_tx, variables=c("ever_lapse_binary")))

# NPDR not on treatment: Moderate
design_npdr_moderate_no_tx <- subset(design, NPDR == "Moderate" & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
data_npdr_moderate_no_tx <- subset(m_data, NPDR == "Moderate" & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
n_npdr_moderate_no_tx <- data_npdr_moderate_no_tx |> nrow()
mfx_npdr_moderate_no_tx <- summary(margins(model, design = design_npdr_moderate_no_tx, data = data_npdr_moderate_no_tx, variables=c("ever_lapse_binary")))

# NPDR  on any treatment: Moderate
design_npdr_moderate_any_tx <- subset(design, NPDR == "Moderate" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
data_npdr_moderate_any_tx <- subset(m_data, NPDR == "Moderate" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
n_npdr_moderate_any_tx <- data_npdr_moderate_any_tx |> nrow()
mfx_npdr_moderate_any_tx <- summary(margins(model, design = design_npdr_moderate_any_tx, data = data_npdr_moderate_any_tx, variables=c("ever_lapse_binary")))

# NPDR not on treatment: Severe
design_npdr_severe_no_tx <- subset(design, NPDR == "Severe" & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
data_npdr_severe_no_tx <- subset(m_data, NPDR == "Severe" & (PRP_flag == 0 & anti_VEGF == 0 & focal_laser_flag == 0 & other_inject == 0))
n_npdr_severe_no_tx <- data_npdr_severe_no_tx |> nrow()
mfx_npdr_severe_no_tx <- summary(margins(model, design = design_npdr_severe_no_tx, data = data_npdr_severe_no_tx, variables=c("ever_lapse_binary")))

# NPDR  on any treatment: Severe
design_npdr_severe_any_tx <- subset(design, NPDR == "Severe" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
data_npdr_severe_any_tx <- subset(m_data, NPDR == "Severe" & (PRP_flag == 1 | anti_VEGF == 1 | focal_laser_flag == 1 | other_inject == 1))
n_npdr_severe_any_tx <- data_npdr_severe_any_tx |> nrow()
mfx_npdr_severe_any_tx <- summary(margins(model, design = design_npdr_severe_any_tx, data = data_npdr_severe_any_tx, variables=c("ever_lapse_binary")))

```

```{r report}

# PDR on any treatment
print("PDR on any treatment:")
print(n_any_tx)
mfx_any_tx |> kable()

# PDR on anti-VEGF only
print("PDR on anti-VEGF only:")
print(n_anti_vegf)
mfx_anti_vegf |> kable()

# PDR on PRP only
print("PDR on PRP only:")
print(n_pdr_prp)
mfx_pdr_prp |> kable()

# NPDR or NoDR not on treatment
print("NPDR or NoDR not on treatment:")
print(n_npdr_nodr_no_tx)
mfx_npdr_nodr_no_tx |> kable()

# NPDR or NoDR any tx
print("NPDR or NoDR any tx:")
print(n_npdr_nodr_any_tx)
mfx_npdr_nodr_any_tx |> kable()

# NPDR not on treatment
print("NPDR not on treatment:")
print(n_npdr_no_tx)
mfx_npdr_no_tx |> kable()

# NPDR  on any treatment
print("NPDR  on any treatment:")
print(n_npdr_any_tx)
mfx_npdr_any_tx |> kable()

# NPDR not on treatment: Mild
print("NPDR not on treatment: Mild:")
print(n_npdr_mild_no_tx)
mfx_npdr_mild_no_tx |> kable()

# NPDR  on any treatment: Mild
print("NPDR  on any treatment: Mild:")
print(n_npdr_mild_any_tx)
mfx_npdr_mild_any_tx |> kable()

# NPDR not on treatment: Moderate
print("NPDR not on treatment: Moderate:")
print(n_npdr_moderate_no_tx)
mfx_npdr_moderate_no_tx |> kable()

# NPDR  on any treatment: Moderate
print("NPDR  on any treatment: Moderate:")
print(n_npdr_moderate_any_tx)
mfx_npdr_moderate_any_tx |> kable()

# NPDR not on treatment: Severe
print("NPDR not on treatment: Severe:")
print(n_npdr_severe_no_tx)
mfx_npdr_severe_no_tx |> kable()

# NPDR  on any treatment: Severe
print("NPDR  on any treatment: Severe:")
print(n_npdr_severe_any_tx)
mfx_npdr_severe_any_tx |> kable()

```